High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-02-04 PNAS_manuscript/PNAS_manuscript.tex
Updates from Overleaf

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$140 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core.
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}.

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of the Lake Superior region showing the distribution of rocks associated with the late Mesoproterozoic Midcontinent Rift. Purple stars mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from  \citealp{Pesonen1983a} are categorized by lithology. All sites from  \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
%\end{figure*}
\end{SCfigure*}

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. B_z denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$;  \citealp{Biggin2014a}) for the anorthosite xenoliths are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a mean of 38.28 $\pm$ 11.92 $\mu$T. This paleointensity value corresponds to a calculated virtual dipole moment of $\sim$89 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S1). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S2;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S2), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (\citealp{Selkin2007a}; Fig. S4).

@@

Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S3). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$140 ZAm$^2$. While the ca. 1140 Abitibi dikes paleointensity estimates do not go to values as low as the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in \cite{Kodama2019}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

@@

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record strong field strengths 1.1 billion years ago. The highest site-level value of the virtual dipole moment would rank in the top 1$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. These high surface field strengths necessitate appreciable power to the late Mesoproterozoic geodynamo.

>>>>

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$82 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$127 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon could challenge the hypothesis of a young inner core. The multiple observed paleointensity transitions from weak to strong in the Paleozoic and the Proterozoic present challenges on the identification of the timing of inner core nucleation based on paleointensity records alone. 
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was followed soon after by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when the inner core unambiguously exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously determine the timing of the initiation of inner core nucleation based on only a few records that span a relatively short time period.

\begin{SCfigure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of the Lake Superior region showing the distribution of rocks associated with the late Mesoproterozoic Midcontinent Rift. Purple stars mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from  \citealp{Pesonen1983a} are categorized by lithology. All sites from  \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{SCfigure*}

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial scale. Note the large contrast in color scale of panels E and F. B$_z$ denotes magnetic field intensity in the vertical direction of the field of view with positive values corresponding to the out of page direction. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

These features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D,F). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of an applied IRM in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM of 300 mT in the orthogonal direction on the Duluth anorthosite, the applied field failed to rotate the remanence directions in plagioclase 1 but resulted in the reversal of remanence in plagioclase 2 which is still oblique to the applied field direction (Fig. S1). In contrast, the remanence magnetizations imaged in the Beaver River anorthosite xenoliths align parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}, S1). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex magma \cite{Miller1990a}.

@@

Following IZZI-style paleointensity experiments \cite{Yu2004a}, 40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity result selection criteria (see Materials and Methods section). 7 anorthosite sites and no diabase site-level results pass these selection criteria. Example NRM/TRM (Arai) plots are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level mean paleointensity values are plotted in Figure \ref{fig:PINT_cooling_corrected} (and provided in Table S1). The paleointensity quality index (Q$_{PI}$; \citealp{Biggin2014a}) for the anorthosite xenoliths are all 5 or 6 (Table S2). The cooling rate-corrected absolute paleointensity estimates from all anorthosite sites have a mean of 38.28 $\pm$ 11.92 $\mu$T. This paleointensity value corresponds to a calculated virtual dipole moment of $\sim$82 ZAm$^2$ ($10^{21} Am^2$) ca. 1092 Ma. All measurement-level paleointensity experiment data are available within the MagIC database (\url{https://earthorg/MagIC/19462/8d3c2258-11ae-4830-b99f-3f6b02eceb7e}; \textit{this private link is provided for the purpose of review; the url will be updated when a doi is generated for this manuscript}).

@@

In addition to estimating paleointensity values by introducing a set of selection criteria to filter our experiment results, we apply an independent method of \citealp{Cych2021a} which uses all experimental data regardless of their NRM/TRM plot statistics to perform bias corrected estimation of paleointensity (BiCEP). This Bayesian probabilistic method is based on an assumption that paleointensity estimates from specimens that come from a same cooling unit are distributed around a true paleointensity value with the various deflections being expressed as the curvature parameter of the NRM/TRM plot \cite{Paterson2011a}. The posterior paleointensity distributions from these sites with high-quality specimen-level data are in agreement with the site-level averages developed using the selection criteria approach (Fig. \ref{fig:PINT_cooling_corrected}; Fig. S2). Overall, the high-quality results from the anorthosite xenoliths of the Beaver River diabase indicate that the anorthosite xenoliths record a high geomagnetic field ca. 1092 Ma.

@@

Rock magnetic data support that the anorthosite specimens that pass the paleointensity selection criteria have dominant magnetic remanence carriers with magnetic properties similar to stochiometric, non-interacting, single domain magnetite, whereas anorthosite samples that failed the paleointensity result selection and all diabase samples have more pronounced populations of non-ideal carriers. Magnetic property measurement system (MPMS) data show both diabase and anorthosite contain (titano)magnetite as evidenced through the presence of the Verwey transition (Fig. S3;  \citealp{Verwey1939a, Feinberg2015a}). Anorthosite specimens from sites that yield successful paleointensity results have Verwey transition temperatures near 120 K as expected for stochiometric magnetite with minimal Ti \cite{Ozdemir1993a}. However, diabase and anorthosite specimens that did not pass our paleointensity selection typically have Verwey transitions that are suppressed toward lower temperatures (Fig. S3), indicating that magnetite grains in the specimens either have relatively higher Ti content or have been partially oxidized \cite{Ozdemir1993a}.

@@

We applied full TRMs to a set of anorthosite specimens with lab fields of 30, 50, 70, and 90 $\mu$T along specimen vertical axes. The results show that this set of anorthosite specimens did not saturate upon the applied fields and there is thus no need for correcting nonlinear thermal remanence acquisitions (\citealp{Selkin2007a}; Fig. S5).

@@

Another consequence of the anorthosite xenoliths having cooled in the interior of thick diabase intrusions is that slow cooling rates can bias paleointensity estimates toward higher values. Large differences in cooling rates between acquisition of an NRM in nature versus a TRM in the lab can result in overestimated paleointensities for single domain grains \cite{Dodson1980a, Halgedahl1980a, Nagy2021a}. From the thermal history model of ref.  \citealp{Zhang2021b}, we can estimate the duration over which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully blocked in their characteristic natural remanence magnetization ($\sim$500\textdegree C;  \citealp{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $\sim1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster through the same temperature interval with an estimated cooling rate of $\sim1.3\times10^{-1}$ $^\circ$C s$^{-1}$. The significant cooling rate difference leads to a predicted $\sim$36\% overestimate of true ancient field following the model of \citealp{Halgedahl1980a} (Fig. S4). This estimate on cooling rate effect is similar to the value of $\sim$35\% overestimate derived from the model of ref. \citealp{Nagy2021a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate-corrected specimen paleointensity estimates together with specimen- and site-level means are shown in Fig. \ref{fig:PINT_cooling_corrected}.

@@

The hypothesis that a weak Ediacaran geomagnetic field is a telltale sign of the lack of an inner core with core nucleation following shortly thereafter may predict that it is the most significant weak to strong field transition in the paleointensity record. However, Fig. \ref{fig:PINT_compilation} shows that transitions from low to high field intensities occurred before, during, and after the Ediacaran. In the Ediacaran record developed to date, there is a two-fold increase in Earth's virtual dipole moment when comparing estimates from the ca. 565 Ma Sept-$\hat{I}$les intrusions \cite{Bono2019a} to those from ca. 550 Ma volcanics of the Skinner Cove Formation \cite{Thallner2021a} (Fig. \ref{fig:PINT_compilation}). In the late Mesoproterozoic, there is at least a six-fold increase within a period of $\sim$35 Myr from a low average virtual dipole moment of $\sim$13 ZAm$^2$ recorded by the ca. 1140 Ma Abitibi dikes which yielded straight Arai plots \cite{Macouin2003a}, to a high moment of $\sim$70 ZAm$^2$ recorded by the ca. 1106 Ma Osler Volcanics, with even stronger values from ca. 1092 Ma by Beaver River anorthosite xenoliths that record virtual dipole moments up to $\sim$127 ZAm$^2$. While the ca. 1140 Abitibi dikes paleointensity estimates do not go to values as low as the ca. 565 Ma Sept-$\hat{I}$les intrusions, this virtual dipole moment increase in the Mesoproterozoic from the ca. 1140 Ma data to the ca. 1100 Ma data is the largest yet documented in the Precambrian on a 10s of millions of years timescale (Fig. \ref{fig:PINT_compilation}). The tempo and scale of this field intensity transition could match with model-based predictions associated with the onset of inner core nucleation \cite{Davies2021a}. This ca. 1.1 Ga timing would be broadly consistent with the ca. 1.3 Ga onset proposed by \citealp{Biggin2015a} albeit later given the exclusion of previous overestimated paleointensity values from the ca. 1.3 Ga Gardar basalts that are superseded by data in \cite{Kodama2019}. However, a model prediction of sustained strong field values following inner core nucleation is challenged by data from the ca. 1070 Ma Bangemall Sills which include a sill with a low virtual dipole moment of $\sim$6.4 ZAm$^2$ (Fig. \ref{fig:PINT_compilation};  \citealp{Lloyd2021b}). Following the Ediacaran, there are also low paleointensity estimates from Devonian rocks such as ca. 370 Ma dikes and lavas of the Siberian Viluy Traps that give virtual dipole moment estimates of 4.3 to 14.9 ZAm$^2$ (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2019a}). These low values as well as data from the ca. 414 Ma Strathmore lava flows \cite{Hawkins2021a}, the ca. 410-380 Ma lava flows of Siberia and the Kola Peninsula \cite{Shcherbakova2017a}, the ca. 408–393 Ma Buribay volcanics \cite{Shcherbakova2021a} and the ca. 332 Ma Kinghorn volcanics \cite{Hawkins2021a} has led to the proposal of this interval as the ``Mid-Paleozoic Dipole Low'' \cite{Hawkins2021a}. This ``Mid-Paleozoic Dipole Low'' is followed by high paleointensity values such that there is a six-fold increase in the database from virtual dipole moment of $\sim$16 ZAm$^2$ at ca. 332 Ma to $\sim$99 ZAm$^2$ ca. 308 Ma in the late Carboniferous (Fig. \ref{fig:PINT_compilation}; \citealp{Hawkins2021a}).

@@

Overall, the high-fidelity paleointensity recorders of the Beaver River anorthosite xenoliths in the well-preserved Midcontinent Rift record strong field strengths 1.1 billion years ago. The highest site-level value of the virtual dipole moment would rank in the top 2$\%$ of those in the database for the Cenozoic Era (the last 66 million years) when there was unequivocally a crystallizing inner core. These high surface field strengths necessitate appreciable power to the late Mesoproterozoic geodynamo.
